---
title: "Interfacing ChemoSpec to PLS"
description: It's easy to connect the two packages
author: "Bryan Hanson"
date: 2021-02-08
categories:
  - R
  - PLS
  - ChemoSpec
bibliography: ../REFS.bib
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, echo = FALSE, results = "hide"}
library("knitr")

set.seed(123)
opts_chunk$set(eval = TRUE, echo = TRUE)
```

The `ChemoSpec` package carries out exploratory data analysis (EDA) on spectroscopic data.  EDA is often described as "letting that data speak", meaning that one studies various descriptive plots, carries out clustering (HCA) as well as dimension reduction (e.g. PCA), with the ultimate goal of finding any natural structure in the data.

As such, `ChemoSpec` does not feature any predictive modeling functions because other packages provide the necessary tools. I do however hear from users several times a year about how to interface a `ChemoSpec` object with these other packages, and it seems like a post about how to do this is overdue.   I'll illustrate how to carry out partial least squares (PLS) using data stored in a `ChemoSpec` object and the package `chemometrics` by Peter Filzmoser and Kurt Varmuza [@chemometrics2017].  One can also use the `pls` package [@PLS2020].

PLS is a technique related to regression and PCA that tries to develop a mathematical model between a matrix of sample vectors, in our case, spectra, and one or more separately measured dependent variables that describe the same samples (typically, chemical analyses).  If one can develop a reliable model, then going forward one can measure the spectrum of a new sample and use the model to predict the value of the dependent variables, presumably saving time and money.  This post will focus on interfacing `ChemoSpec` objects with the needed functions in `chemometrics`.  I won't cover how to evaluate and refine your model, but you can find plenty on this in @Filzmoser2009 chapter 4, along with further background (there's a lot of math in there, but if you aren't too keen on the math, gloss over it to get the other nuggets). Alternatively, take a look at the vignette that ships with `chemometrics` via `browseVignettes("chemometrics")`.

As our example we'll use the marzipan NIR data set that one can download in Matlab format from [here](http://www.models.life.ku.dk/Marzipan).[^1]  The corresponding publication is [@Christensen2004].  This data set contains NIR spectra of marzipan candies made with different recipes and recorded using several different instruments, along with data about moisture and sugar content.  We'll use the data recorded on the NIRSystems 6500 instrument, covering the 400-2500 nm range.  The following code chunk gives a summary of the data set and shows a plot of the data. Because we are focused on how to carry out PLS, we won't worry about whether this data needs to be normalized or otherwise pre-processed (see the Christensen paper for lots of details).

```{r showSpectra}
library("ChemoSpec")
load("Marzipan.RData")
sumSpectra(Marzipan)
plotSpectra(Marzipan, which = 1:32, lab.pos = 3000)
```

In order to carry out PLS, one needs to provide a matrix of spectroscopic data, with samples in rows (let's call it $X$, you'll see why in a moment).  Fortunately this data is available directly from the `ChemoSpec` object as `Marzipan$data`.[^2]  One also needs to provide a matrix of the additional dependent data (let's call it $Y$).  It is critical that the order of rows in $Y$ correspond to the order of rows in the matrix of spectroscopic data, $X$.

Since we are working in `R` we know there are a lot of ways to do most tasks.  Likely you will have the additional data in a spreadsheet, so let's see how to bring that into the workspace.  You'll need samples in rows, and variables in columns.  For your sanity and error-avoidance, you should include a header of variable names and the names of the samples in the first column.  Save the spreadsheet as a csv file.  I did these steps using the sugar and moisture data from the original paper.  Read the file in as follows.

```{r readCSV}
Y <- read.csv("Marzipan.csv", header = TRUE)
str(Y)
```

The function we'll be using wants a matrix as input, so convert the data frame that `read.csv` generates to a matrix.  Note that we'll select only the numeric variables on the fly, as unlike a data frame, a matrix can only be composed of one data type.

```{r toMatrix}
Y <- as.matrix(Y[, c("sugar", "moisture")])
str(Y)
```

Now we are ready to carry out PLS.  Since we have a multivariate $Y$, we need to use the appropriate function (use `pls1_nipals` if your $Y$ matrix is univariate).

```{r PLS}
library("chemometrics")
pls_out <- pls2_nipals(X = Marzipan$data, Y, a = 5)
```

And we're done!  Be sure to take a look at `str(pls_out)` to see what you got back from the calculation.  For the next steps in evaluating your model, see section 3.3 in the `chemometrics` vignette.


[^1]: I have converted the data from Matlab to a `ChemoSpec` object; if anyone wants to know how to do this let me know and I'll put up a post on that process.
[^2]: `str(Marzipan)` will show you the structure of the `ChemoSpec` object (or in general, any `R` object).  The official definition of a `ChemoSpec` object can be seen via `?Spectra`.
