---
title: "Metabolic Phenotyping Protocol Part 3"
description: "Implementing the Statistical Analysis in Metabolic Phenotyping Protocol of Blaise *et al.*"
date: "2022-05-01"
categories: [R, ChemoSpec, Metabolomics, PLS, PLS-DA]
bibliography:
  - REFS.bib
  - grateful-refs.bib
---

```{r SetUp, echo = FALSE, eval = TRUE, results = "hide", warning = FALSE, message = FALSE}

# R options & configuration:
set.seed(123)
rm(list = ls())
suppressPackageStartupMessages(library("ChemoSpec"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("grateful"))

descCS <- packageDescription("ChemoSpec")
descCSU <- packageDescription("ChemoSpecUtils")

options(ChemoSpecGraphics = "ggplot2")

# Stuff specifically for knitr:

# Hook for figure size control
knitr::opts_hooks$set(sq.fig = function(options) {

  if (isFALSE(options$sq.fig)) {
    # custom fig dimensions given, use w/o further delay
    if ((!is.null(options$fig.width)) & (!is.null(options$fig.height))) return(options)
    # otherwise set the default wide aspect ratio
    if ((is.null(options$fig.width)) & (is.null(options$fig.height))) {
      options$fig.width = 6
      options$fig.height = 3.5
    }
  }

  if (isTRUE(options$sq.fig)) {
    options$fig.width = 5
    options$fig.height = 5
  }
  options
})

# choices here are designed to work with the hook
knitr::opts_chunk$set(fig.align = "center", sq.fig = FALSE, tidy = TRUE,
  fig.width = NULL, fig.height = NULL, out.width = "80%")
```

* Part 1 of this series is [here](https://chemospec.org/posts/2022-02-01-Protocol-Pt1/).
* Part 2 of this series is [here](https://chemospec.org/posts/2022-03-24-Protocol-Pt2/).

If you aren't familiar with `ChemoSpec`, you might wish to look at the introductory [vignette](https://bryanhanson.github.io/ChemoSpec/articles/ChemoSpec.html) first.

*In this series of posts we are following the protocol as described in the printed publication closely [@Blaise2021].  The authors have also provided a [Jupyter notebook](https://github.com/Gscorreia89/chemometrics-tutorials).  This is well worth your time, even if Python is not your preferred language, as there are additional examples and discussion for study.*

# Read in the Data

Load the `Spectra` object we created in Part 2 so we can summarize it.

```{r getData}
library("ChemoSpec")
load("Worms2.RData") # restores the "Worms2" Spectra object
sumSpectra(Worms2)
```

# Exploratory Data Analysis, Con't.

If you recall in Part 2 we removed five samples.  Let's re-run PCA without these samples and show the key plots.  We will simply report these here without much discussion; they are pretty much as expected.

```{r pca}
c_pca <- c_pcaSpectra(Worms2, choice = "autoscale")
```

```{r}
#| label: fig-screeAlt
#| fig-cap: "Scree plot (recommended style)."
plotScree(c_pca)
```

```{r}
#| label: fig-scores12
#| sq.fig: true
#| fig-cap: "Score plot for PCs 1 and 2. Compare to protocol figure 7d."
p <- plotScores(Worms2, c_pca, pcs = 1:2, ellipse = "rob", tol = 0.02)
p
```

```{r}
#| label: fig-scores23
#| sq.fig: true
#| fig-cap: "Score plot for PCS 2 and 3."
p <- plotScores(Worms2, c_pca, pcs = 2:3, ellipse = "rob", leg.loc = "bottomleft", tol = 0.02)
p
```

One thing the published protocol does not explicitly discuss is an inspection of the loadings, but it is covered in the [Jupyter notebook](https://github.com/Gscorreia89/chemometrics-tutorials).  The loadings are useful in order to see if any particular frequencies are driving the separation of the samples in the score plot. Let's plot the loadings (@fig-loadings). Remember that these data were autoscaled, and hence all frequencies, including noisy frequencies, will contribute to the separation.  If we had not scaled the data, these plots would look dramatically different.

```{r}
#| label: fig-loadings
#| sq.fig: true
#| fig-cap: "Loadings for PC1 and PC2."
p <- plotLoadings(Worms2, c_pca, loads = 1:2)
p
```

The s-plot is another very useful way to find peaks that are important in separating the samples (@fig-splot); we can see that the peaks around $\delta$ 1.30-1.32, 1.47-1.48, and 3.03-3.07 are important drivers of the separation in the score plot.  Having discovered this, one can investigate the source of those peaks.

```{r}
#| label: fig-splot
#| sq.fig: true
#| fig-cap: "s-Plot for PC1."

p <- sPlotSpectra(Worms2, c_pca, tol = 0.001)
p
```

## Supervised Analysis with PLS-DA

`ChemoSpec` carries out exploratory data analysis, which is an unsupervised process.  The next step in the protocol is PLS-DA (partial least squares - discriminant analysis). I have written about `ChemoSpec` + PLS [here](https://chemospec.org/posts/2021-02-08-PLS/2021-02-08-PLS.html) if you would like more background on plain PLS. However, PLS-DA is a technique that combines data reduction/variable selection along with classification. We'll need the `mixOmics` package (@mixOmics) package for this analysis; note that loading it replaces the `plotLoadings` function from `ChemoSpec`.

```{r}
library("mixOmics")
```

@fig-plsda-indiv shows the score plot; the results suggest that classification and modeling may be successful.  The `splsda` function carries out a single *sparse* computation. One computation should not be considered the ideal answer; a better approach is to use cross-validation, for instance the `bootsPLS` function in the `bootsPLS` package (@bootsPLS which uses `splsda` under the hood).  However, that computation is too time-consuming to demonstrate here.


```{r}
X <- Worms2$data
Y <- Worms2$groups
splsda <- splsda(X, Y, ncomp = 8)
```

```{r}
#| label: fig-plsda-indiv
#| fig-cap: "sPLS-DA plot showing classification."
#| sq.fig: true
#| tidy: false
plotIndiv(splsda,
  col.per.group = c("#FB0D16FF", "#FFC0CBFF", "#511CFCFF", "#2E94E9FF"),
  title = "sPLS-DA Score Plot", legend = TRUE, ellipse = TRUE)
```

To estimate the number of components needed, the `perf` function can be used. The results are in @fig-plsda-perf and suggest that five components are sufficient to describe the data.

```{r}
#| label: fig-plsda-perf
#| fig-cap: "Evaluation of the PLS-DA performance."
#| sq.fig: true
perf.splsda <- perf(splsda, folds = 5, nrepeat = 5)
plot(perf.splsda)
```

At this point, we have several ideas of how to proceed.  Going forward, one might choose to focus on accurate classification, or on determining which frequencies should be included in a predictive model.  Any model will need to refined and more details extracted.  The reader is referred to the case study from the [mixOmics](http://mixomics.org/case-studies/splsda-srbct/) folks which covers these tasks and explains the process.

```{r}
#| eval: false
#| results: "hide"
#| warning: false
#| echo: false
get_pkgs_info(pkgs = c("mixOmics", "bootsPLS"))
```

---

This post was created using `ChemoSpec` version `r descCS$Version` and `ChemoSpecUtils` version `r descCSU$Version`.
