---
title: "Earth's Field NMR"
description: "What can we measure?  Why or Why Not?"
date: "2023-06-26"
categories: [NMR, EF-NMR]
format:
  html
bibliography: references.bib
draft: true
---

```{r}
#| label: setup
#| echo: false
#| results: hide
library("latex2exp")
```

Let's take a closer look from first principles what kinds of information one can glean from EF-NMR.  We'll restrict our discussion to spin $\frac{1}{2}$ nuclei with ~100% abundance, like $\mathrm{^{1}H}$, $\mathrm{^{31}P}$ or $\mathrm{^{19}F}$ -- you'll see why soon enough.  Here are some relevant physical parameters for these nuclei.

| Nuclei | Mag. Moment | Gyromagnetic ratio $\gamma$ | Larmor Freq. |
| ------ | -----------: | --------: | ------------: |
| $\mathrm{^{1}H}$ | 2.7928 | 26.7522 | 100 |
| $\mathrm{^{19}F}$ | 2.6269 | 25.1815 | 94 |
| $\mathrm{^{31}P}$ | 1.136 | 10.8394 | 40.5 |

: Important NMR parameters for ~100% abundant spin $\frac{1}{2}$  nuclei.  Magnetic moment is in units of ??.  The units of $\gamma$ are $10^7$ rad $T^{-1}$ $s^{-1}$.  Larmor frequency is relative. {#tbl-params}

Excellent general references on NMR theory are and .

## EF-NMR Signals are in the Audio Range

What is the Larmor (resonance) frequency in earth's field? Earth's magnetic field [varies](https://en.wikipedia.org/wiki/Earth%27s_magnetic_field) from about 25 to 65 $\mu$T; we'll use an intermediate value of 45 $\mu$T for our calculations.  The Larmor frequency is given by the equation:

$$
\nu_{L} = \lvert\frac{\gamma}{2\pi}\rvert B_o
$$

Notice there is a simple linear relation between $\nu_{L}$ and $B_o$.[^1]  If we plug in values for our nuclei we get the following values in Hz:

```{r, echo = FALSE}
gammaS <- c(26.7522, 25.1815, 10.8394) * 10e7
names(gammaS) <- c("1H", "19F", "31P")
EF <- 45e-6
nuL <- gammaS * EF/(2*pi)
nuL
```

What we have shown here is that for EF-NMR, resonance frequencies are in the [audio](https://en.wikipedia.org/wiki/Audio_frequency) (20 - 20,000 Hz) and lower [radio](https://en.wikipedia.org/wiki/Radio_frequency) (20,000 Hz +) frequency range.  Why is this important?  It greatly simplifies signal detection because audio receivers are essentially radios, and the electronics for working in this frequency range are extremely well worked out, and not expensive to buy or build.

## Line Widths are Very Narrow

## No Chemical Shift Information

One of the characteristics of high-field NMR which makes it so useful is the dispersion of chemical shifts as a function of structure.  Unfortunately, EF-NMR has effectively zero chemical shift dispersion.  The equation for computing chemical shift, $\delta$, is:

$$
\delta = \frac{\nu_{sample} - \nu_{reference}}{\nu_{B_o}}
$$

where the units are:

$$
ppm = \frac{Hz}{MHz}
$$

since $\delta$ is a field strength independent quantity.  Taking $\nu_{reference}$ to be zero, e.g. TMS added to the sample, we can rearrange the equation to get $\nu_{sample}$.  Consider the compound $\mathrm{CH_{3}Br}$ whose methyl group has a chemical shift of 2.63 ppm.  Using an earth's field $^{1}H$ Larmor frequency of 19.1 KHz, we can compute the shift of $\mathrm{CH_{3}Br}$ in Hz as 0.0191 Hz.  This is an extremely small value, smaller than the typical linewidth in earth's field. For further comparison, we can do the same calculation for $\mathrm{CH_{2}Br_{2}}$ which has a shift of 4.90 ppm.  The result is exactly the same, 0.0191 Hz.  We can see that these two compounds with differing numbers of halogens, which would be trivial to distinguish with a low field bench-top instrument operating at 80 MHz, are indistinguishable at earth's field.  This is partly the result of the chemical shift range for $^{1}H$ being fairly small at 0-15 ppm, but more so due to the very small value of earth's field.

## Heteronuclear Couplings

## Opportunity: 

Basic NMR theory tells us that the energy difference between the two quantum states for a spin $\frac{1}{2}$ nucleus is proportional to the field strength $B_o$:

$$
\Delta E = \gamma \hbar B_o
$$

where $\hbar$ is $\frac{h}{2\pi}$. A plot for $\mathrm{^{1}H}$ is shown in @fig-DeltaE; the right-most point corresponds to a 1,000 MHz instrument.  Clearly as $B_o$ goes to zero the $\Delta E$ goes to zero.

```{r, echo = FALSE, results = "hide"}
library("ggplot2")
```

```{r, echo = FALSE}
planck <- 6.626e-34
hbar <- planck/(2 * pi)
gamma <- 26.7522
B0 <- seq(0, 23.49,length.out = 20)
larmor <- gamma * B0 / (2 * pi)
DeltaE <- gamma * hbar * B0
DF <- data.frame(B0, lower = -0.5*DeltaE, upper = 0.5*DeltaE)
```

```{r}
#| label: fig-DeltaE
#| fig.height: 3
#| echo: false
#| fig.cap: "$\\Delta$E as a function of field strength $B_o$"
p <- ggplot(data = DF)
p <- p + geom_line(aes(B0, upper)) + geom_line(aes(B0, lower))
p <- p + geom_point(aes(B0, upper)) + geom_point(aes(B0, lower))
p <- p + labs(x = "Field Strength, Tesla", y = TeX("$\\Delta$E, Joules"))
p
```

[^1]: The gyromagnetic ratio can be negative, hence the absolute value is taken here.
